[Re] Predicting metabolic biomarkers of human inborn errors of metabolism

### 2018-02-14 article/Mondeel_Ogundipe_Westerhoff-2017.tex
RE: Federico
- Discussion of the flux variability cobrapy function.
- bold math notation in the definition of the FVA problem (changed the ReScience template to include the bm package)
- consistent definition of exchange reactions in the notebooks
- Some references to the manuscript in the notebook comments.
- extra discussion of phi = 0 or phi > 0
- Redraw the Fig. 1 network and include the fluxes

\documentclass[10pt,a4paper,onecolumn]{article}
% \usepackage[utf8]{inputenc}
\usepackage{marginnote}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{authblk,etoolbox}
\usepackage{titlesec}
\usepackage{calc}
\usepackage{hyperref}
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor=
{
      Thierry D.G.A. Mondeel,
      Vivian Ogundipe,
      Hans V. Westerhoff,
  },
            pdftitle=
{
[Re] Predicting metabolic biomarkers of human inborn errors of metabolism
},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=blue,
            pdfborder={0 0 0}}
\urlstyle{same}
\usepackage{tcolorbox}
\usepackage{ragged2e}
\usepackage{fontspec}
\usepackage{fontawesome}
\usepackage{caption}
\usepackage{listings}
\lstnewenvironment{code}{\lstset{language=Haskell,basicstyle=\small\ttfamily}}{}

@@

Each biomarker will be associated with a prediction for being either
elevated or decreased in the mutant case, as comparted to the wild type.
A biomarker is considered to have an increased, or reduced,
extracellular concentration in the mutant case if, when plotting the
wild-type and mutant flux-variability intervals on a horizontal axis
(see below), both the lower and upper boundary of the mutant interval
are shifted to the right, or left, respectively. If the two borders of
the mutant interval move in opposite directions with respect to the
wild-type interval, the result is scored as `unchanged' and the boundary
metabolite is not considered to be a biomarker.

@@

\begin{align}
\text{Maximize } &\pm v_i, \text{ such that for all k} \nonumber \\
S v &= 0 \nonumber \\
Z  &\geq \phi Z_{\text{max}} \nonumber \\
\alpha_k \leq & ~ v_k \leq \beta_k 
\end{align}

Here \(v\) is the column vector of fluxes representing all reactions in
the model, \(v_i\) is the flux of the exchange reaction of a given
boundary metabolite \(i\), \(\alpha_k\) is the, possibly negative, lower
bound for reaction \(k\) and similarly \(\beta_k\) is the, possibly
negative, upper bound for reaction \(k\). The reaction bounds are the
\(V_{max}\)'s of the reactions and are part of the metabolic network
definition, typically they are set to either \(0\) or \(\pm 1000\) if
the true, biological, \(V_{max}\)'s are not known. These values allow
specification of irreversible reactions by setting the lower bound (or
upper bound) to zero. The index \(i\) specifies a single reaction that
is to be maximized, whereas the index \(k\) is used to index the bound
constraints that exist for each reaction, including \(v_i\). \(v_i\) may
be maximized in the forward (positive flux) or the reverse (negative
flux) direction if allowed by the bounds on \(v_i\). This is indicated
by the \(\pm\) symbol. \(S\) is the stoichiometry matrix for the
metabolic network of size \(mxr\) where \(m\) is the number of
metabolites and \(r\) the number of reactions. The numeric, typically
integer, elements of \(S\) represent the stoichiometry coefficients of
each metabolite \(m\) in each reaction \(r\). Consequently, for large
networks, this matrix is sparse (has many zero entries). \(Z = c^T ~ v\)
is the objective function defined for the map, entailing a linear
combination, defined by column vector \(c\), of one or more reactions.
The vector \(c\) is an indicator of objective reactions, i.e.~it
contains a value of \(1\) in rows corresponding to fluxes that are to be
included in the objective, and zeros in all other rows. Due to the
vector multiplication \(c^T ~ v\), the objective \(Z\) sums the fluxes
of the reactions that correspond to rows containing a value of \(1\) in
\(c\). \(Z_{\text{max}}\) denotes the maximal value of the objective
function. We define \(0 \leq \phi \leq 1\) to be an arbitrary minimal
fraction of the maximal objective function value that has to be
achieved. When \(\phi=1\) we ask for the range of flux allowed for
reaction \(v_i\) while maintaining the maximal value of the linear
combination of objective fluxes. When setting \(\phi = 0\), one is
asking for the allowable flux range through reaction \(v_i\) regardless
of the value of the objective function (although the sign has to be
maintained). The latter, i.e. \(\phi=0\), is the case considered in the
method by Shlomi et al.

@@

After the loop is completed, contradictory predictions for \(m\) between
the affected reactions \(r\) are dealt with based on a majority rule.
When there is an equal number of elevated and reduced predictions, the
boundary metabolite is considered to be unchanged, i.e.~not a biomarker.

@@

\begin{figure}
\centering
\includegraphics{Figures/scheme_shlomi2009_figure1}
\caption{A metabolic network used to illustrate the predictions of
biomarkers of IEMs as originally considered in the work by Shlomi et al.
for their prediction of biomarkers of IEMs. All metabolites but M3 have
an exchange reaction allowing for uptake or secretion of the
metabolites. M2 synthesis is fully dependent on uptake of M1 and M5 or
M6. The IEM considered is one that leads to deletion of the enzyme
converting M1 plus M3 to M2 plus M4.}\label{fig:Shlomi2009_figure_1}
\end{figure}

@@

The network in Figure ~\ref{fig:Shlomi2009_figure_1} consists of seven
metabolites, \(6\) of which (i.e.~all except M3) have exchange
reactions. We consider a hypothetical deletion of the enzyme catalyzing
the conversion of M1 into M2 coupled to the conversion of M3 into M4.
Shlomi et al. graphically provide the biomarker prediction for this
network in their Figure 1A. Here, we do so numerically and this paper is
accompanied by a Jupyter notebook reproducing the biomarker prediction.
In our analysis, all exchange reactions were given inward bounds of
\(-10\) and outward bounds of \(+1000\) so that the internal reaction
bounds (of \(\pm 1000\)) are never reached. Our Table
~\ref{tbl:toy_model_results} shows that all exchangeable metabolites
except M5 and M7 are predicted to be robust biomarkers.

@@

\hypertarget{tbl:toy_model_results}{}
\begin{longtable}[]{@{}lllc@{}}
\caption{\label{tbl:toy_model_results}Biomarker predictions using our
implementation with respect to the mutation affecting the conversion of
M1 plus M3 to M2 plus M4 in the network of Figure
~\ref{fig:Shlomi2009_figure_1}. H.C. stands for `highly confident'
indicating, as defined by Shlomi et al., no overlap between the WT and
mutant intervals. Between brackets {[}{]} we report the lower and upper
bounds of the FVA intervals we obtained. }\tabularnewline
\toprule
ID & Prediction & WT & Mutant\tabularnewline
\midrule
\endfirsthead
\toprule
ID & Prediction & WT & Mutant\tabularnewline
\midrule
\endhead
M1 & H.C. Elevated & {[}-10.0, -1.0{]} & {[}0.0, 0.0{]}\tabularnewline
M2 & H.C. Reduced & {[}1.0, 10.0{]} & {[}0.0, 0.0{]}\tabularnewline
M4 & Reduced & {[}1.0, 20.0{]} & {[}0.0, 10.0{]}\tabularnewline
M6 & Elevated & {[}-10.0, 9.0{]} & {[}0.0, 10.0{]}\tabularnewline
M7 & Unchanged & {[}-10.0, 0.0{]} & {[}-10.0, 0.0{]}\tabularnewline
M5 & Unchanged & {[}-10.0, 0.0{]} & {[}-10.0, 0.0{]}\tabularnewline
\bottomrule
\end{longtable}

>>>>

\documentclass[10pt,a4paper,onecolumn]{article}
% \usepackage[utf8]{inputenc}
\usepackage{marginnote}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{authblk,etoolbox}
\usepackage{titlesec}
\usepackage{calc}
\usepackage{hyperref}
\usepackage{bm}
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor=
{
      Thierry D.G.A. Mondeel,
      Vivian Ogundipe,
      Hans V. Westerhoff,
  },
            pdftitle=
{
[Re] Predicting metabolic biomarkers of human inborn errors of metabolism
},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=blue,
            pdfborder={0 0 0}}
\urlstyle{same}
\usepackage{tcolorbox}
\usepackage{ragged2e}
\usepackage{fontspec}
\usepackage{fontawesome}
\usepackage{caption}
\usepackage{listings}
\lstnewenvironment{code}{\lstset{language=Haskell,basicstyle=\small\ttfamily}}{}

@@

Each biomarker will be associated with a prediction for being either
elevated or decreased in the mutant case, as comparted to the wild type.
A biomarker is considered to have an increased, or reduced,
extracellular concentration in the mutant case if, when plotting the
wild-type and mutant flux-variability intervals on a horizontal axis,
both the lower and upper boundary of the mutant interval are shifted to
the right, or left, respectively (see Figures
~\ref{fig:Shlomi2009_figure_1}B and ~\ref{fig:Shlomi2009_figure_3}B). If
the two borders of the mutant interval move in opposite directions with
respect to the wild-type interval, the result is scored as `unchanged'
and the boundary metabolite is not considered to be a biomarker.

@@

\begin{align}
\text{Maximize } &\pm v_i, \text{ such that for all } k \nonumber \\
\bm{S} \bm{v} &= \bm{0} \nonumber \\
Z  &\geq \phi Z_{\text{max}} \nonumber \\
\alpha_k \leq & ~ v_k \leq \beta_k 
\end{align}

Here \(\bm{v}\) is the column vector of fluxes representing all
reactions in the model, \(v_i\) is the flux of the exchange reaction of
a given boundary metabolite \(i\), \(\alpha_k\) is the, possibly
negative, lower bound for reaction \(k\) and similarly \(\beta_k\) is
the, possibly negative, upper bound for reaction \(k\). The reaction
bounds are the \(V_{max}\)'s of the reactions and are part of the
metabolic network definition, typically they are set to either \(0\) or
\(\pm 1000\) if the true, biological, \(V_{max}\)'s are not known. These
values allow specification of irreversible reactions by setting the
lower bound (or upper bound) to zero. The index \(i\) specifies a single
reaction that is to be maximized, whereas the index \(k\) is used to
index the bound constraints that exist for each reaction, including
\(v_i\). \(v_i\) may be maximized in the forward (positive flux) or the
reverse (negative flux) direction if allowed by the bounds on \(v_i\).
This is indicated by the \(\pm\) symbol. \(\bm{S}\) is the stoichiometry
matrix for the metabolic network of size \(m \times r\) where \(m\) is
the number of metabolites and \(r\) the number of reactions. The
numeric, typically integer, elements of \(\bm{S}\) represent the
stoichiometry coefficients of each metabolite \(m\) in each reaction
\(r\). Consequently, for large networks, this matrix is sparse (has many
zero entries). \(Z = \bm{c}^T \bm{v}\) is the objective function defined
for the map, entailing a linear combination, defined by column vector
\(\bm{c}\), of one or more reactions. The vector \(\bm{c}\) is an
indicator of objective reactions, i.e.~it contains a value of \(1\) in
rows corresponding to fluxes that are to be included in the objective,
and zeros in all other rows. Due to the vector multiplication
\(\bm{c}^T \bm{v}\), the objective \(Z\) sums the fluxes of the
reactions that correspond to rows containing a value of \(1\) in
\(\bm{c}\). \(Z_{\text{max}}\) denotes the maximal value of the
objective function. We define \(0 \leq \phi \leq 1\) to be an arbitrary
minimal fraction of the maximal objective function value that has to be
achieved. When \(\phi=1\) we ask for the range of flux allowed for
reaction \(v_i\) while maintaining the maximal value of the linear
combination of objective fluxes. When setting \(\phi = 0\), one is
asking for the allowable flux range through reaction \(v_i\) regardless
of the value of the objective function (although the sign has to be
maintained). The latter, i.e. \(\phi=0\), is the case considered in the
method by Shlomi et al. The choice of \(\phi\) matters because when
\(phi \neq 0\) we are requiring flux to flow through the set of
objective reactions. This may entail a forced directionality through
reactions that are required to ultimately produce flux in the objective
reactions. These additional limitation in freedom of the flux pattern
may subsequently impact the biomarker predictions. The approach
discussed here entails predictions purely based on the network topology
and therefore sets \(\phi = 0\).

@@

After the loop (steps 1-5) is completed, contradictory predictions for
\(m\) between the affected reactions \(r\) are dealt with based on a
majority rule. When there is an equal number of elevated and reduced
predictions, the boundary metabolite is considered to be unchanged,
i.e.~not a biomarker.

The flux variability analyses that yield the exchange flux intervals
computed in step 1, 2 and 4 are handled by the
flux\_analysis.variability.flux\_variability\_analysis function in
Cobrapy. The input for this function comprises the model, along with its
associated flux bounds which change during step 1, 2 and 4, the list of
reactions for which to calculate the flux intervals and a fraction of
the optimal objective flux to minimally obtain (fraction\_of\_optimum).
In our case we pass the model with flux bounds as described above, lists
of exchange reactions for the metabolites of interest and a fraction of
optimum of zero.

@@

\begin{figure}
\centering
\includegraphics{Figures/Reproduction_Shlomi_fig_1}
\caption{(A) A metabolic network used to illustrate the predictions of
biomarkers of IEMs as originally considered in the work by Shlomi et al.
for their prediction of biomarkers of IEMs. All metabolites but M3 have
an exchange reaction allowing for uptake or secretion of the
metabolites. M2 synthesis is fully dependent on uptake of M1 and M5 or
M6. The IEM considered is one that leads to deletion of the enzyme
converting M1 plus M3 to M2 plus M4. (B) Visual illustration of the
predicted exchange intervals for each of the metabolites with an
exchange reaction. The wild-type (WT) interval is indicated in black,
the mutant interval in red. Each interval is associated with a numeric
interval predicted by the biomarker prediction algorithm and with a
qualitative prediction for the level of the metabolite in the biofluids
(see methods). H.C. is an abbreviation for a highly confident biomarker,
see main text.}\label{fig:Shlomi2009_figure_1}
\end{figure}

@@

The network in Figure ~\ref{fig:Shlomi2009_figure_1}A consists of seven
metabolites, \(6\) of which (i.e.~all except M3) have exchange
reactions. We consider a hypothetical deletion of the enzyme catalyzing
the conversion of M1 into M2 coupled to the conversion of M3 into M4.
Shlomi et al. graphically provide the biomarker prediction for this
network in their Figure 1A. Here, we do so graphically and numerically
in Figure ~\ref{fig:Shlomi2009_figure_1}B and this paper is accompanied
by a Jupyter notebook reproducing the biomarker prediction. In our
analysis, all exchange reactions were given inward bounds of \(-10\) and
outward bounds of \(+1000\) so that the internal reaction bounds (of
\(\pm 1000\)) are never reached. Our Table ~\ref{tbl:toy_model_results}
shows that all exchangeable metabolites except M5 and M7 are predicted
to be robust biomarkers.

@@
